Applications of the cross-entropy method in diffusion processes

### 2013-12-24 writedown/first.hitting/cross_entropy/cross_entropy.tex
revised by Wei

\begin{document}
\title{Applications of the cross-entropy method in diffusion processes}
%\author{Wei Zhang}
\author{Wei Zhang\textsuperscript{2}\and Han Wang\textsuperscript{1} \and
  Carsten Hartmann\textsuperscript{1,*}\and 
Marcus Weber\textsuperscript{2}\and Christof Sch\"{u}tte\textsuperscript{1,2}}
\footnotetext[1]{Institute of Mathematics, Freie Universit\"{a}t Berlin, Arnimallee 6, 14195 Berlin, Germany}
\footnotetext[2]{Zuse Institute Berlin, Takustrasse 7, 14195 Berlin, Germany}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Corresponding author. Email: chartman@mi.fu-berlin.de}
\date{}
\maketitle
\begin{abstract}
  The cross-entropy method in diffusion process setting is studied. The
  general algorithm is proposed and is applied to design efficient importance sampling for rare events as well as to solve optimal
  control problem. Several examples are studied numerically.
\end{abstract}
\begin{keywords}
importance sampling, optimal control, cross-entropy method, Monte Carlo
method, rare event, change of measure. 
\end{keywords}
\begin{AMS}
\end{AMS}
\section{Introduction}
\label{intro}
Diffusion process has been extensively investigated because of its wide
applications in mathematics, physics, economics, and so on.
We are mainly interested in two topics in diffusion process, i.e. importance sampling and optimal control. 
For importance sampling, it is one kind of variance reduction technique 
in Monte Carlo simulations and becomes especially necessary in rare event simulation.
For optimal control, the purpose is to driven the dynamical system in certain
optimal way and has strong application backgrounds in finance, climate
science, mechanics \cite{fleming2006}. We emphasize that these two different areas are actually related and can
be unified as the same problem, i.e. how to change the measure in certain optimal way.

Here we take an applied point of view, i.e. we aim at designing efficient
algorithms to find the way to do the importance sampling, or to find the
optimal control strategy. One main reason that hinder efficient algorithms is
due to the high dimensionality of the dynamical systems, as in the case of
climate system, biological system and so on. However, when the dynamical
systems have multiscale feature, efficient algorithms may be designed by
taking this feature into account. See \cite{ip-dupuis-multiscale, ip-eric,
zhws13} on importance sampling and \cite{control_schuette, zlph2013} on
optimal control for more details.

In this work, we consider these two topics together as designing the change of
measure in an optimal way. If we further assume the existence of the optimal
change of measure $\mu^*$, the problem becomes to find approximation of
$\mu^*$ under some measure distance.

On the other hand, the Kullback-Leibler divergence, although not a metric, is
widely used to quantified the similarity of two probability measures. 
What is more, given certain target probability measure $\mu^*$, the so-called cross-entropy method has been
developed recently and provides a general algorithm to find the minimizer of the
Kullback-Leibler divergence w.r.t. $\mu^*$, among a family of parameterized
probability measures. It has wide applications in different areas, such as
Monte Carlo simulation, optimization, and machine learning \cite{ce_book}. The
main purpose of this paper is to formulate the cross-entropy method in the
context of diffusion process, i.e. for measures defined in path space, and
discuss how to apply this method in importance sampling and optimal control
problems.

This paper is organized as follows. In Section~\ref{sec-om-functional},
using Girsanov theorem, we derive the Onsager-Machlup functional for a given trajectory.
In Section~\ref{sec-ce-is}, the cross-entropy method in path space is
formulated for the importance sampling problem.
In Section~\ref{sec-ocp}, by using the dual relationship between importance
sampling and optimal control problem, we discuss the cross-entropy method for
optimal control problem. The cross-entropy algorithm is summarized in
Section~\ref{sec-ce-algo}. Several numerical examples are studied in
Section~\ref{sec-examples}. Some discussions are presented in
Section~\ref{sec-discuss}.

@@

We could also rewrite it as Stratonovich integral with the relationship

@@

      I_1(z) = \frac{1}{2} \int_0^\tau \Big|\frac{dz}{ds} - b(z_s)\Big|_a^2 ds

@@

      trajectory space, i.e. $\forall z = (x,y) $ is a trajectory, we have 
      $d\mu = \exp(-\frac{\beta}{2} I_1(z)) d\nu$ (\textbf{or formally :}  $\mu(dz) =
      \exp(-\frac{\beta}{2} I_1(z)) \nu(dy)$, which is better ?), where $\nu$ is Wiener

@@

  Since $I_2$ is simply the Stratonovich version of $I_1$, we use $I_1$ for
  simplicity.
  Starting from dynamics
  d z_s &= \widetilde{b}(z_s) ds + \sqrt{2\beta^{-1}} d\widetilde{w}_s, \quad s >0  \\
  z_0&=z
\label{dynamics-11}
the above argument can be repeated to get 
  d\widetilde{\mu} &=\exp\Big(- \frac{\beta}{4}\int_0^\tau \Big|\frac{dz}{ds} -
  \widetilde{b}(z_s)\Big|^2 ds\Big) d\nu
  \label{measure-2}
Divided it by (\ref{measure-1}), we have
  \frac{d\widetilde{\mu}}{d\mu} = \exp\Big(- \frac{\beta}{4}\Big[2\int_0^\tau (dz_s,
  b(z_s) - \widetilde{b}(z_s)) + \int_0^\tau (|\widetilde{b}(z_s)|^2 - |b(z_s)|^2) ds\Big]
  \Big)
Then we substitute the dynamics (\ref{dynamics-1}) in above (in principle we
can also use dynamics (\ref{dynamics-11}), but $\widetilde{w}$ is not Brownian motion
under $\mu$ ...), after
simplification, we have 
  \frac{d\widetilde{\mu}}{d\mu} = \exp\Big(-
  \frac{\beta^{1/2}}{\sqrt{2}}\int_0^\tau (dw_s,
b(z_s) - \widetilde{b}(z_s)) - \frac{\beta}{4}\int_0^\tau (|\widetilde{b}(z_s)
- b(z_s)|^2) ds\Big)
Let $u_s = \frac{1}{\sqrt{2}} (b(z_s) - \widetilde{b}(z_s))$, it gives
  \frac{d\widetilde{\mu}}{d\mu} = \exp\Big(- \beta^{1/2}\int_0^\tau (dw_s, u_s)
) - \frac{\beta}{2}\int_0^\tau |u_s|^2 ds\Big)
So we recover the more general situation in our IS paper for $\sigma =
\sqrt{2}$.
\section{Cross entropy method for importance sampling}

@@

In this section, we discuss how to apply the cross entropy

@@

Let $\mathcal{D}$ be certain subset in the path space and
$\mathds{1}_{\mathcal{D}}$ be the
characteristic function on $\mathcal{D}$. 
In the following, for simplicity, we will use $z$ to denote a path trajectory
$\{z_s\}$ and keep the notation $z_s \in \mathbb{R}^n$ for the state at time $s$. 
Let $F(\cdot)$ be certain path functional and will be written as $F(z)$ for
path $z$. 
We consider Monte Carlo method with importance sampling to compute the quantity
\begin{align}
  l = \mathrm{E}(F(z)~|~z_0 = x)
  \label{mean-l}
\end{align}
where we assume $F(\cdot)$ has the following type
\begin{align}
  F(z) = \mathds{1}_{\mathcal{D}}(z) \quad \mbox{or} \quad F(z) = \exp(-\beta\int_0^\tau G(z_s) \,ds).
  \label{path-fun-f}
\end{align}
\begin{remark}
  A special and interesting case is when $\mathcal{D} = \{z, \tau(z) \le \lambda\}$,
  $\lambda > 0$ is fixed. Then $l = \mathrm{E}(\mathds{1}_{\mathcal{D}}(z)) =
  \mathcal{P}(\tau \le \lambda)$, i.e. the probability of the trajectories
  stopping before time $\lambda$. This quantity may provide us some details of the
  distribution of the stopping time $\tau$ and help us understand the
  behaviors of the dynamics.
\end{remark}

We suppose $m$ basis functions on state space $\{\phi_i\}_{1\le i \le m}$ are
available, where $\phi_i : \mathbb{R}^n\rightarrow \mathbb{R}$. 
We will add a control force to the original dynamics (\ref{dynamics-1})
and the importance sampling will be realized by simulating trajectories from
this modified dynamics. Specifically, given $\bm{w} = (w_1, w_2,
\cdots, w_m)^T \in \mathbb{R}^m$,
the corresponding control force is state-dependent and is supposed to be 
\begin{align}
  u(x; \bm{w}) = \sum\limits_{i = 1}^{m} w_i \nabla \phi_i(x), \quad x \in
  \mathbb{R}^n\,.
  \label{control-force}
\end{align}
Therefore, for fixed parameter $\bm{w}$, the controlled dynamics is 
\begin{align}
\begin{split}
  d z_s &= b(z_s) ds + \sqrt{2}u(z_s;\bm{w}) ds + \sqrt{2\beta^{-1}} dw_s, \quad t \le s \le \tau \\
     z_0 &=z
  \end{split}
\label{dynamics-2}
\end{align}
Let $\mu_{\bm{w}}$ be the probability measure of dynamics (\ref{dynamics-2})
w.r.t. $\bm{w}$. 
From Section \ref{sec-om-functional},  we know $\mu_{\bm{w}}$ has density 
$f(z;\bm{w}) = \exp(-\frac{\beta}{2} I(z ; \bm{w}))$ w.r.t. Wiener
measure $\nu$, with
\begin{align}
  I(z ; \bm{w}) = \frac{1}{2}\int_0^\tau \Big|\frac{dz}{ds}- b(z_s) -
  \sqrt{2}\sum\limits_{i = 1}^{m} w_i \nabla \phi_i(z_s)\Big|^2 ds + \beta^{-1}
  \int_0^\tau \operatorname{div}\, \Big(b(z_s) + \sqrt{2}\sum\limits_{i =
  1}^{m} w_i \nabla \phi_i(z_s)\Big) ds
  \label{om-fun-w}
\end{align}
and the original dynamics (\ref{dynamics-1}) corresponds to $\bm{w} = \bm{0}$ (we use $\mu = \mu_{\bm{0}}$).
Given a probability measure $\mu_g$ on path space with $d\mu_g = g(z)d\nu$, 
we can design a unbiased Monte Carlo estimator with importance sampling method as 
\begin{align}
  l_N = \frac{1}{N} \sum_{i=1}^{N}  \frac{F(z_i)f(z_i;\bm{0})}{g(z_i)}
\end{align}
where trajectories $z_i, i = 1, \cdots, N$ are sampled from measure $\mu_g$.
The optimal measure $\mu_{g^*}$, which gives
zero-variance estimator, should satisfy 
\begin{align}
  g^*(z) = \frac{F(z)f(z;\bm{0})}{l}, \hspace{3mm} \forall z
  \label{optimal_pdf}
\end{align}
However, $g^*(z)$ depends on the path $z$ and is not easy to compute.

The idea of cross entropy method is to find the best approximation of $g^*(z)$ among
$f(z;\bm{w}), \bm{w} \in \mathbb{R}^m$. More specifically, define the

@@

  \min_{\bm{w} \in \mathbb{R}^m} D(\mu_{g^*}, \mu_{\bm{w}}).

@@

  \max_{\bm{w} \in \mathbb{R}^m} \mathrm{E}_{\mu} ( F(z) \log f(z;\bm{w}))
  = \max_{\bm{w} \in \mathbb{R}^m} \mathrm{E}_{\mu_{\bm{v}}} ( F(z)
  w(z;\bm{v}) \log f(z;\bm{w})), \hspace{3mm} \forall \bm{v} \in

@@

  w(z;\bm{v}) = \frac{f(z;\bm{0})}{f(z;\bm{v})}.

@@

\max_{\bm{w} \in \mathbb{R}^m} \frac{1}{N} \sum_{i=1}^{N} F(z_i) w(z_i, \bm{v}) \log f(z_i;\bm{w}) 

@@

  \frac{1}{N} \sum_{i=1}^{N} F(z_i) w(z_i, \bm{v}) \nabla_{\bm{w}} \log
  f(z_i;\bm{w}) = 0.

@@

as linear equations $A\bm{w} = \bm{r}$, where $\forall 1\le i,j\le m$, 

@@

  A_{ij} &= \frac{1}{N}\sum_{k = 1}^{N} F(z_k)w(z_k, \bm{v}) \int_0^\tau
    \nabla\phi_i(z_k)\nabla\phi_j(z_k)\,ds, \\
     r_i &= \frac{1}{\sqrt{2}N}\sum_{k = 1}^{N} F(z_k)w(z_k, \bm{v})
    \Big[\int_0^\tau \Big(\frac{dz_k}{ds} - b(z_s)\Big)\nabla\phi_i(z_k)\,ds - \beta^{-1} \int_0^\tau \operatorname{div} \nabla \phi_i(z_k)\,ds\Big]

@@

  defined similarly. The aim is to find the optimal $\bm{w} \in
  \mathbb{R}^m$ and the corresponding measure $\mu^\pi_{\bm{w}}$ to do

@@

	dy_s & = b_2(x_s, y_s)\, ds + \sqrt{2} u(x_s, y_s ; \bm{w})\, ds + \sqrt{2\beta^{-1}} dw_s

@@

      with $u(x_s, y_s ; \bm{w}) = \sum\limits_{i=1}^{m} w_i \nabla \phi_i(x_s,
      y_s)$ and optimize the coefficients $\bm{w}$. All the derivations are

@@

\section{Cross entropy method for optimal control problem}

@@

In this section, we consider applying the cross entropy method to solve optimal control problem

@@

Consider the optimal control problem with cost function
\begin{equation}
  J(u) = \mathrm{E}_{\mu_u}\left[\int_0^\tau \left(G(z_s) + \frac{1}{2} |u(z_s,s)|^2
\right) ds ~\bigg|~ z_0 = x\right].
\end{equation}
where $G : \mathbb{R}^n\rightarrow \mathbb{R}$ is nonnegative, $u$ is the
control force and $\mathrm{E}_{\mu_u}$ denotes the expectation under probability measure $\mu_u$,
which is generated by the controlled dynamics
of (\ref{dynamics-1-sigma}), namely,
\begin{align}
 \begin{split}
   d z_s &= b(z_s) ds + \sqrt{2}\sigma(z_s) u(z_s, s) ds + \sqrt{2\beta^{-1}} \sigma(z_s)\,dw_s, \quad 0 < s < \tau \\
    z_0&=x
  \end{split}
  \label{dynamics-1-sigmma-c}
\end{align}

Let $\hat{u} = \argmin J(u)$. For simplicity, we denote $\mu, \hat{\mu}$ as
the probability measure corresponding to $u \equiv 0$ and $\hat{u}$ respectively. Using the dual relation
    \begin{align}
      -\beta^{-1}\log \mathrm{E}_\mu\left[\exp\Big(-\beta\int_0^\tau G(z_s)\, ds\Big)\right] = \inf_u J(u) = J(\hat{u})
      \label{dual-relation}
    \end{align}
    and Jensen's inequality, we know that 
    \begin{align}
    \exp\Big(-\beta\int_0^\tau G(z_s)\,ds\Big)\frac{d\mu}{d\hat{\mu}} =
    \mathrm{E}_\mu\left[\exp\Big(-\beta\int_0^\tau G(z_s)\, ds\Big)\right],
    \hspace{3mm} \hat{\mu}-a.s.
    \label{jensen-as}
  \end{align}
From the above we have 
  \begin{align}
    J(u) &= \mathrm{E}_{\hat{\mu}} \left\{\Big[\int_0^\tau \Big(G(z_s) + \frac{1}{2} |u(z_s,
    s)|^2\Big)\,ds\Big]\frac{d\mu_u}{d\hat{\mu}}\right\}  \\
    & = J(\hat{u}) + \mathrm{E}_{\hat{\mu}}\Big[\Big(\beta^{-1}\log
    \frac{d\mu}{d\hat{\mu}} + \int_0^\tau \frac{1}{2} |u(z_s,
    s)|^2\,ds\Big)\frac{d\mu_u}{d\hat{\mu}}\Big]
\end{align}
After simplification, we get the following simple relationship 
\begin{align}
J(u) = J(\hat{u}) + \beta^{-1}D(\mu_u|\, \hat{\mu}).
\label{J-entropy}
\end{align}
where $D(\cdot, \cdot)$ is the Kullback-Leibler divergence in (\ref{kl-dist}).

We are interested in finding the optimal control $\hat{u}$. For high
dimensional dynamics ($n$ is large), this becomes difficult.  
Similarly as the in previous section, we suppose the control force can be written
as (\ref{control-force}) with $m$ predefined basis functions $\{\phi_i\}_{1\le
i \le m}$. And our problem is simplified to solve
\begin{align}
  \min_{\bm{w} \in \mathbb{R}^m} J(u(z;\bm{w})).
  \label{J-w}
\end{align}
under dynamics (\ref{dynamics-1-sigmma-c}).

From (\ref{J-entropy}), we know that solving (\ref{J-w})
is equivalent to minimize the Kullback-Leibler divergence $D$ between
$\mu_u$ and $\hat{\mu}$, which, however, is still not easy. On the other hand, inspired by 
the discussions in Section \ref{sec-ce-is}, cross entropy method can be
applied to minimize $D(\hat{\mu} \,|\, \mu_u)$ (notice this is different with
(\ref{J-w}), because the Kullback-Leibler divergence is not symmetric). Following
the notations in Section \ref{sec-ce-is} and from (\ref{jensen-as}), we know the optimal measure
satisfies
\begin{align}
  g^*(z) \propto \exp\Big(-\beta\int_0^\tau G(z_s) \,ds\Big) f(z ; 0),
  \quad \forall z\,,
\end{align}
therefore the problem is reduced to the case in Section \ref{sec-ce-is} with $F(z) = \exp(-\beta\int_0^\tau G(z_s) \,ds)$.
\section{Cross entropy algorithm}
\label{sec-ce-algo}
In this section, we consider how to implement the cross entropy algorithm in
order to maximize (\ref{ce-2-discrete}), which is equivalent to (\ref{kl-dist-d}) in the discrete case. 
We suppose the optimal density $g^*(z)$ is given by (\ref{optimal_pdf}) and
$F(z)$ takes the forms in (\ref{path-fun-f}).

>>>>

\begin{document}
\title{Applications of the cross-entropy method in diffusion processes}
%\author{Wei Zhang}
\author{Wei Zhang\textsuperscript{2}\and Han Wang\textsuperscript{1} \and
  Carsten Hartmann\textsuperscript{1,*}\and 
Marcus Weber\textsuperscript{2}\and Christof Sch\"{u}tte\textsuperscript{1,2}}
\footnotetext[1]{Institute of Mathematics, Freie Universit\"{a}t Berlin, Arnimallee 6, 14195 Berlin, Germany}
\footnotetext[2]{Zuse Institute Berlin, Takustrasse 7, 14195 Berlin, Germany}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Corresponding author. Email: chartman@mi.fu-berlin.de}
\date{}
\maketitle
\begin{abstract}
  The cross-entropy method in the diffusion process setting is studied. The
  general algorithm is proposed and is applied to design efficient importance sampling for rare events as well as to solve optimal
  control problem of diffusion process. Several examples are studied
  numerically to demonstrate this method.
\end{abstract}
\begin{keywords}
importance sampling, optimal control, cross-entropy method, Monte Carlo
method, rare event, change of measure. 
\end{keywords}
\begin{AMS}
\end{AMS}
\section{Introduction}
\label{intro}
Diffusion process has been extensively investigated because of its wide
applications in mathematics, physics, economics, and so on.
Here we are mainly interested in two topics in diffusion process, i.e. importance sampling and optimal control. 
For importance sampling, it is one kind of variance reduction technique 
in Monte Carlo simulations and becomes especially necessary in rare event simulation.
For optimal control, the purpose is to driven the dynamical system in certain
optimal way and has strong applicational backgrounds in finance, climate
science, mechanics \cite{fleming2006}. We emphasize that these two different areas are actually related and can
be unified as the same problem, i.e. how to change the measure in certain optimal way.

Here we take an applied point of view, i.e. we aim at designing efficient
algorithms to find the way to do the importance sampling, or to find the
optimal control strategy. One main reason that designing efficient algorithms
is urgent and challenging is due to the high dimensionality of the dynamical
systems, as in the case of climate system, biological system and so on. However, when the dynamical
systems have multiscale feature, efficient algorithms may be designed by
taking this feature into account. See \cite{ip-dupuis-multiscale, ip-eric,
zhws13} on importance sampling and \cite{control_schuette, zlph2013} on
optimal control for more details.

In this work, we consider these two topics together as how to design the change of
measure in an optimal way. If we further assume the existence of the optimal
change of measure $\mu^*$, the problem can be roughly described as how to find the approximation of
$\mu^*$ under some measure distance. For the choice of measure distances, we will use the Kullback-Leibler
divergence, which, although not a metric, is
widely used to quantified the similarity of two probability measures. 
What is more, given certain target probability measure $\mu^*$, the so-called cross-entropy method has been
developed recently and provides a general algorithm to find the minimizer of the
Kullback-Leibler divergence w.r.t. $\mu^*$, among a family of parameterized
probability measures. It has wide applications in different areas, such as
Monte Carlo simulation, optimization, and machine learning \cite{ce_tutorial, ce_book}. The
main purpose of this paper is to formulate the cross-entropy method in the
context of diffusion process, i.e. for measures defined in path space, and
discuss the applications of this method in importance sampling and optimal control
problems.

This paper is organized as follows. In Section~\ref{sec-om-functional},
using Girsanov theorem, we derive the Onsager-Machlup functional for a given trajectory.
In Section~\ref{sec-ce-is}, the cross-entropy method in path space is
formulated for the importance sampling problem.
In Section~\ref{sec-ocp}, by using the dual relationship between importance
sampling and optimal control problem, we discuss the cross-entropy method for
optimal control problem. The cross-entropy algorithm is summarized in
Section~\ref{sec-ce-algo}. Several numerical examples are studied in
Section~\ref{sec-examples}. Discussions are presented in
Section~\ref{sec-discuss}.

@@

We could also rewrite it as Stratonovich integral using the relationship

@@

      I_1(z) = \frac{1}{2} \int_0^\tau \Big|\frac{dz}{ds} -
      b(z_s)\Big|_{a(z)}^2 ds

@@

      \textbf{trajectory space}, i.e. $\forall z = (x,y) $ is a trajectory, we have 
      $d\mu = \exp(-\frac{\beta}{2} I_1(z)) d\nu$, where $\nu$ is Wiener

@@

\section{Cross-entropy method for importance sampling}

@@

In this section, we discuss how to apply the cross-entropy

@@

Let $\mathcal{D}$ be certain subset in the path space and
$\textbf{1}_{\mathcal{D}}$ be the
characteristic function on $\mathcal{D}$. 
In the following, for simplicity, we will use $z$ to denote a path trajectory
$\{z_s\}$ and keep the notation $z_s \in \mathbb{R}^n$ for the state at time $s$. 
Let $F(\cdot)$ be certain path functional and will be written as $F(z)$ for
path $z$. 
We consider Monte Carlo method with importance sampling to compute the quantity
\begin{align}
  l = \mathrm{E}(F(z)~|~z_0 = x)
  \label{mean-l}
\end{align}
where we assume $F(\cdot)$ has the following type
\begin{align}
  F(z) = \textbf{1}_{\mathcal{D}}(z) \quad \mbox{or} \quad F(z) = \exp(-\beta\int_0^\tau G(z_s) \,ds).
  \label{path-fun-f}
\end{align}
  A special and interesting case is when $\mathcal{D} = \{z, \tau(z) \le \lambda\}$,
  $\lambda > 0$ is fixed. Then $l = \mathrm{E}(\textbf{1}_{\mathcal{D}}(z)) =
  \mathcal{P}(\tau \le \lambda)$, i.e. the probability of the trajectories
  stop before time $\lambda$. This quantity may provide us some details of the
  distribution of the stopping time $\tau$ and help us understand the
  behaviors of the dynamics.

We suppose $m$ basis functions on state space $\{\phi_i\}_{1\le i \le m}$ are
available, where $\phi_i : \mathbb{R}^n\rightarrow \mathbb{R}$. 
We will add a control force to the original dynamics (\ref{dynamics-1})
and the importance sampling will be realized by simulating trajectories from
this modified dynamics. Specifically, given $\bm{\omega} = (\omega_1, \omega_2,
\cdots, \omega_m)^T \in \mathbb{R}^m$,
the corresponding control force is state-dependent and is supposed to be 
\begin{align}
  u(x; \bm{\omega}) = \sum\limits_{i = 1}^{m} \omega_i \nabla \phi_i(x), \quad x \in
  \mathbb{R}^n\,.
  \label{control-force}
\end{align}
Therefore, for fixed parameter $\bm{\omega}$, the controlled dynamics is 
\begin{align}
\begin{split}
  d z_s &= b(z_s) ds + \sqrt{2}u(z_s;\bm{\omega}) ds + \sqrt{2\beta^{-1}} dw_s, \quad t \le s \le \tau \\
     z_0 &=z
  \end{split}
\label{dynamics-2}
\end{align}
Let $\mu_{\bm{\omega}}$ be the probability measure of dynamics (\ref{dynamics-2}). 
From Section \ref{sec-om-functional},  we know $\mu_{\bm{\omega}}$ has density 
$f(z;\bm{\omega}) = \exp(-\frac{\beta}{2} I(z ; \bm{\omega}))$ w.r.t. Wiener
measure $\nu$, with
\begin{align}
  I(z ; \bm{\omega}) = \frac{1}{2}\int_0^\tau \Big|\frac{dz}{ds}- b(z_s) -
  \sqrt{2}\sum\limits_{i = 1}^{m} \omega_i \nabla \phi_i(z_s)\Big|^2 ds + \beta^{-1}
  \int_0^\tau \operatorname{div}\, \Big(b(z_s) + \sqrt{2}\sum\limits_{i =
  1}^{m} \omega_i \nabla \phi_i(z_s)\Big) ds
  \label{om-fun-w}
\end{align}
and the original dynamics (\ref{dynamics-1}) corresponds to $\bm{\omega} = \bm{0}$ (we use $\mu = \mu_{\bm{0}}$).
Given a probability measure $\mu_g$ on path space with $d\mu_g = g(z)d\nu$, 
we can design a unbiased Monte Carlo estimator for (\ref{mean-l}) as 
\begin{align}
  l_N = \frac{1}{N} \sum_{i=1}^{N}  \frac{F(z_i)f(z_i;\bm{0})}{g(z_i)}
\end{align}
where trajectories $z_i, i = 1, \cdots, N$ are sampled from measure $\mu_g$.
The optimal measure $\mu_{g^*}$, which gives
zero-variance estimator, should satisfy 
\begin{align}
  g^*(z) = \frac{F(z)f(z;\bm{0})}{l}, \hspace{3mm} \forall z
  \label{optimal_pdf}
\end{align}
However, $g^*(z)$ depends on the path $z$ and is not easy to compute.

The idea of cross-entropy method is to find the best approximation of $g^*(z)$ among
$f(z;\bm{\omega}), \bm{\omega} \in \mathbb{R}^m$. More specifically, define the

@@

  \min_{\bm{\omega} \in \mathbb{R}^m} D(\mu_{g^*}, \mu_{\bm{\omega}}).

@@

  \max_{\bm{\omega} \in \mathbb{R}^m} \mathrm{E}_{\mu} ( F(z) \log
  f(z;\bm{\omega}))
  = \max_{\bm{\omega} \in \mathbb{R}^m} \mathrm{E}_{\mu_{\bm{v}}} ( F(z)
  \eta(z;\bm{v}) \log f(z;\bm{\omega})), \hspace{3mm} \forall \bm{v} \in

@@

  \eta(z;\bm{v}) = \frac{f(z;\bm{0})}{f(z;\bm{v})}.

@@

\max_{\bm{\omega} \in \mathbb{R}^m} \frac{1}{N} \sum_{i=1}^{N} F(z_i)
\eta(z_i; \bm{v}) \log f(z_i;\bm{\omega}) 

@@

  \frac{1}{N} \sum_{i=1}^{N} F(z_i) \eta(z_i; \bm{v}) \nabla_{\bm{\omega}} \log
  f(z_i;\bm{\omega}) = 0.

@@

as linear equations $A\bm{\omega} = \bm{r}$, where $\forall 1\le i,j\le m$, 

@@

  A_{ij} &= \frac{1}{N}\sum_{k = 1}^{N} F(z_k)\eta(z_k; \bm{v}) \int_0^\tau
  \nabla\phi_i(z_{k,s})\nabla\phi_j(z_{k,s})\,ds, \\
     r_i &= \frac{1}{\sqrt{2}N}\sum_{k = 1}^{N} F(z_k)\eta(z_k; \bm{v})
  \Big[\int_0^\tau \Big(\frac{dz_k}{ds} - b(z_s)\Big)\nabla\phi_i(z_{k,s})\,ds
  - \beta^{-1} \int_0^\tau \operatorname{div} \nabla \phi_i(z_{k,s})\,ds\Big]

@@

  defined similarly. The aim is to find the optimal $\bm{\omega} \in
  \mathbb{R}^m$ and the corresponding measure $\mu^\pi_{\bm{\omega}}$ to do

@@

	dy_s & = b_2(x_s, y_s)\, ds + \sqrt{2} u(x_s, y_s ; \bm{\omega})\, ds + \sqrt{2\beta^{-1}} dw_s

@@

      with $u(x_s, y_s ; \bm{\omega}) = \sum\limits_{i=1}^{m} \omega_i \nabla \phi_i(x_s,
      y_s)$ and optimize the coefficients $\bm{\omega}$. All the derivations are

@@

\section{Cross-entropy method for optimal control problem}

@@

In this section, we consider applying the cross-entropy method to solve optimal control problem

@@

Consider the optimal control problem with cost
function \cite{control_schuette,zlph2013}
\begin{equation}
  J(u) = \mathrm{E}_{\mu_u}\left[\int_0^\tau \left(G(z_s) + \frac{1}{2} |u(z_s,s)|^2
\right) ds ~\bigg|~ z_0 = x\right],
\end{equation}
where $G : \mathbb{R}^n\rightarrow \mathbb{R}$ is nonnegative, $u$ is the
control force and $\mathrm{E}_{\mu_u}$ denotes the expectation under probability measure $\mu_u$,
which is generated by the controlled dynamics
of (\ref{dynamics-1-sigma}), namely,
\begin{align}
 \begin{split}
   d z_s &= b(z_s) ds + \sqrt{2}\sigma(z_s) u(z_s, s) ds + \sqrt{2\beta^{-1}} \sigma(z_s)\,dw_s, \quad 0 < s < \tau \\
    z_0&=x
  \end{split}
  \label{dynamics-1-sigmma-c}
\end{align}

Suppose the lower bound of $J(u)$ can be achieved and let $\hat{u} = \argmin J(u)$. For simplicity, we denote $\mu, \hat{\mu}$ as
the probability measure corresponding to $u \equiv 0$ and $\hat{u}$
respectively. Using the dual relation \cite{DaiPra1996, zlph2013}
    \begin{align}
      -\beta^{-1}\log \mathrm{E}_\mu\left[\exp\Big(-\beta\int_0^\tau G(z_s)\, ds\Big)\right] = \inf_u J(u) = J(\hat{u})
      \label{dual-relation}
    \end{align}
    and Jensen's inequality, we know that 
    \begin{align}
    \exp\Big(-\beta\int_0^\tau G(z_s)\,ds\Big)\frac{d\mu}{d\hat{\mu}} =
    \mathrm{E}_\mu\left[\exp\Big(-\beta\int_0^\tau G(z_s)\, ds\Big)\right],
    \hspace{3mm} \hat{\mu}-a.s.
    \label{jensen-as}
  \end{align}
From the above we have 
  \begin{align}
    J(u) &= \mathrm{E}_{\hat{\mu}} \left\{\Big[\int_0^\tau \Big(G(z_s) + \frac{1}{2} |u(z_s,
    s)|^2\Big)\,ds\Big]\frac{d\mu_u}{d\hat{\mu}}\right\}  \\
    & = J(\hat{u}) + \mathrm{E}_{\hat{\mu}}\Big[\Big(\beta^{-1}\log
    \frac{d\mu}{d\hat{\mu}} + \int_0^\tau \frac{1}{2} |u(z_s,
    s)|^2\,ds\Big)\frac{d\mu_u}{d\hat{\mu}}\Big]
\end{align}
After simplification, we get the following simple relationship 
\begin{align}
J(u) = J(\hat{u}) + \beta^{-1}D(\mu_u|\, \hat{\mu}).
\label{J-entropy}
\end{align}
where $D(\cdot, \cdot)$ is the Kullback-Leibler divergence in (\ref{kl-dist}).

We are interested in finding the optimal control $\hat{u}$. For high
dimensional dynamics ($n$ is large), this becomes difficult.  
Similarly as the in previous section, we suppose the control force can be written
as (\ref{control-force}) with $m$ predefined basis functions $\{\phi_i\}_{1\le
i \le m}$. And our problem is simplified to solve
\begin{align}
  \min_{\bm{\omega} \in \mathbb{R}^m} J(u(z;\bm{\omega})).
  \label{J-w}
\end{align}
under dynamics (\ref{dynamics-1-sigmma-c}).

From (\ref{J-entropy}), we know that solving (\ref{J-w})
is equivalent to minimize the Kullback-Leibler divergence $D$ between
$\mu_u$ and $\hat{\mu}$, which, however, is still not easy. On the other hand, inspired by 
the discussions in Section \ref{sec-ce-is}, we can apply cross-entropy method 
to minimize $D(\hat{\mu} \,|\, \mu_u)$ (notice this is different from
(\ref{J-w}), because the Kullback-Leibler divergence is not symmetric), and
hope that (\ref{J-w}) can be solved approximately. Following
the notations in Section \ref{sec-ce-is} and from (\ref{jensen-as}), we know the optimal measure
satisfies
\begin{align}
  g^*(z) \propto \exp\Big(-\beta\int_0^\tau G(z_s) \,ds\Big) f(z ; 0),
  \quad \forall z\,,
\end{align}
therefore the problem is reduced to the case in Section \ref{sec-ce-is} with $F(z) = \exp(-\beta\int_0^\tau G(z_s) \,ds)$.
\section{Cross-entropy algorithm}
\label{sec-ce-algo}
In this section, we consider how to implement the cross-entropy algorithm in
order to maximize (\ref{ce-2-discrete}), which is equivalent to (\ref{kl-dist-d}) in the discrete case. 
We suppose the optimal density $g^*(z)$ is given by (\ref{optimal_pdf}) and
$F(z)$ takes the forms in (\ref{path-fun-f}).
